\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-09-06 paper.qmd
Intro paragraph for PBMC3k
PBMC3k intro

In the field of single-cell studies, a common analytical task involves clustering to identify groups of cells with similar expression profiles. NLDR methods are commonly used to display clusters, and help to verify the results. For example, @chen2023 illustrates the use of UMAP to identify clusters in Human Peripheral Blood Mononuclear Cells (PBMC3k). There are $2622$ single cells. First 9 principal components are used to generate the UMAP. @fig-NLDR-variety (a) is the reproduction of the published plot. The objective is to select a reasonable UMAP layout, representing the non-linear structure of the digit 1 dataset in $10\text{-}D$.

@@

To determine whether the UMAP representation with the specific (hyper-)parameter choice suggested by @chen2023 preserves the data structure present in $9\text{-}D$ (@fig-model-pbmc-author-proj (b)), we visualize the model constructed with UMAP overlaid on the $9\text{-}D$ data. The @fig-NLDR-variety (a) shows three well-separated clusters with big separations. However, as shown in @fig-model-pbmc-author-proj, there is no big separation between three clusters in $9\text{-}D$. Therefore, the suggested UMAP representation (@fig-NLDR-variety (a)) does not accurately represent the structure of PBMC3k dataset in $9\text{-}D$. But, when visualizing the *model-in-the-data-space* some unobserved structures can be seen. Some clusters have non-linear continuity patterns and high-density patches (@fig-model-pbmc-author-proj).

>>>>

In the field of single-cell studies, a common analytical task involves clustering to identify groups of cells with similar expression profiles. NLDR methods are commonly used to display clusters, and help to verify the results. For example, @chen2023 illustrates the use of UMAP to identify clusters in Human Peripheral Blood Mononuclear Cells (PBMC3k). There are $2622$ single cells. First 9 principal components are used to generate the UMAP. @fig-NLDR-variety (a) is the reproduction of the published plot. The objective is to assess the published layout, and if it does not accurately represent the three clusters with small separations of the PBMC3k dataset (@fig-model-pbmc-author-proj (a2)), then select a reasonable \gD{} layout.

@@

To determine whether the UMAP representation with the specific (hyper-)parameter choice suggested by @chen2023 preserves the data structure present in $9\text{-}D$ (@fig-model-pbmc-author-proj (b)), we visualize the model constructed with UMAP overlaid on the $9\text{-}D$ data. The @fig-NLDR-variety (a) shows three well-separated clusters with big separations. However, as shown in @fig-model-pbmc-author-proj, there is no big separation between three clusters in $9\text{-}D$. Therefore, the suggested UMAP representation (@fig-NLDR-variety (a)) does not accurately represent the structure of PBMC3k dataset in $9\text{-}D$.

<!-- But, when visualizing the *model-in-the-data-space* some unobserved structures can be seen. Some clusters have non-linear continuity patterns and high-density patches (@fig-model-pbmc-author-proj). -->

### 2024-09-06 paper.tex
PBMC3k intro

In the field of single-cell studies, a common analytical task involves
clustering to identify groups of cells with similar expression profiles.
NLDR methods are commonly used to display clusters, and help to verify
the results. For example, \citet{chen2023} illustrates the use of UMAP
to identify clusters in Human Peripheral Blood Mononuclear Cells
(PBMC3k). There are \(2622\) single cells. First 9 principal components
are used to generate the UMAP. Figure~\ref{fig-NLDR-variety} (a) is the
reproduction of the published plot. The objective is to select a
reasonable UMAP layout, representing the non-linear structure of the
digit 1 dataset in \(10\text{-}D\).

To determine whether the UMAP representation with the specific
(hyper-)parameter choice suggested by \citet{chen2023} preserves the
data structure present in \(9\text{-}D\)
(Figure~\ref{fig-model-pbmc-author-proj} (b)), we visualize the model
constructed with UMAP overlaid on the \(9\text{-}D\) data. The
Figure~\ref{fig-NLDR-variety} (a) shows three well-separated clusters
with big separations. However, as shown in
Figure~\ref{fig-model-pbmc-author-proj}, there is no big separation
between three clusters in \(9\text{-}D\). Therefore, the suggested UMAP
representation (Figure~\ref{fig-NLDR-variety} (a)) does not accurately
represent the structure of PBMC3k dataset in \(9\text{-}D\). But, when
visualizing the \emph{model-in-the-data-space} some unobserved
structures can be seen. Some clusters have non-linear continuity
patterns and high-density patches
(Figure~\ref{fig-model-pbmc-author-proj}).

>>>>

In the field of single-cell studies, a common analytical task involves
clustering to identify groups of cells with similar expression profiles.
NLDR methods are commonly used to display clusters, and help to verify
the results. For example, \citet{chen2023} illustrates the use of UMAP
to identify clusters in Human Peripheral Blood Mononuclear Cells
(PBMC3k). There are \(2622\) single cells. First 9 principal components
are used to generate the UMAP. Figure~\ref{fig-NLDR-variety} (a) is the
reproduction of the published plot. The objective is to assess the
published layout, and if it does not accurately represent the three
clusters with small separations of the PBMC3k dataset
(Figure~\ref{fig-model-pbmc-author-proj} (a2)), then select a reasonable
\gD{} layout.

To determine whether the UMAP representation with the specific
(hyper-)parameter choice suggested by \citet{chen2023} preserves the
data structure present in \(9\text{-}D\)
(Figure~\ref{fig-model-pbmc-author-proj} (b)), we visualize the model
constructed with UMAP overlaid on the \(9\text{-}D\) data. The
Figure~\ref{fig-NLDR-variety} (a) shows three well-separated clusters
with big separations. However, as shown in
Figure~\ref{fig-model-pbmc-author-proj}, there is no big separation
between three clusters in \(9\text{-}D\). Therefore, the suggested UMAP
representation (Figure~\ref{fig-NLDR-variety} (a)) does not accurately
represent the structure of PBMC3k dataset in \(9\text{-}D\).
